Discrete Quantum Theories and Computing

### 2018-03-03 dissertation.tex
dissertation.lyx "2.1.1 The single qubit problem" Updating...

A convenience for computing probability and relative state properties
is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\frac{\theta}{2}}$, that
is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\frac{\theta}{2}}\alpha_{0},\rme^{\rmi\frac{\theta}{2}}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007},
which we can write down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\frac{\theta}{2}}$
(``fibering out by the circle $\Sphere{1}$''). The standard form
of this maps (``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Vec{X}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $\left(\phi,\psi\right)$ on
the surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

\begin{figure}[tbh]
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.48\columnwidth]{hopfS2}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.48\columnwidth]{hopfFibers}
\par\end{centering}
}\hfill{} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $\rme^{\rmi\frac{\theta}{2}}$), each corresponding
to a single point on the sphere in (a).\protect \\
\todo{{Since }the Hopf fibration{ is fibering out by} the circle
$\Sphere{1}${ }on the three-sphere $\Sphere{3}$ embedded in $\R^{4}$.
Need to confirm whether (b) is a figure in $\R^{4}$ and how to understand
the next sentence.}\protect \\
Points in (a) are color coded corresponding to circles in (b), e.g.,
one pole contains the red elliptical circle that would become an infinite-radius
circle in a slightly different projection, and the opposite pole corresponds
to the large perfectly round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

@@

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Vec{a}$ and $\Vec{b}$ on $\Sphere{2}$,
the shortest rotation carrying the unit normal $\Hat{a}$ to the unit
normal $\Hat{b}$ is the SLERP (spherical linear interpolation) 
\begin{equation}
S(\Hat{a},\Hat{b},t)=\Hat{a}\,\frac{\sin((1-t)\theta)}{\sin\theta}\,+\,\Hat{b}\,\frac{\sin(t\theta)}{\sin\theta}\ ,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\theta$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Fig.~\ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

>>>>

A convenience for computing probability and relative state properties
is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\frac{\theta}{2}}$, that
is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\frac{\theta}{2}}\alpha_{0},\rme^{\rmi\frac{\theta}{2}}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007,wiki:HopfFibration},
which we can write down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\frac{\theta}{2}}$
(``fibering out by the circle $\Sphere{1}$''). The standard form
of this maps (``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{eq:HopfFibration}\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Hat{a}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\|\Hat{a}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $\left(\phi,\psi\right)$ on
the surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

\begin{figure}[tbh]
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.48\columnwidth]{hopfS2}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.48\columnwidth]{hopfFibers}
\par\end{centering}
}\hfill{} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration, \ref{eq:HopfFibration},
as a family of circles (the paths of $\rme^{\rmi\frac{\theta}{2}}$)
on the three-sphere $\Sphere{3}$. Although $\Sphere{3}$ cannot be
directly embedded in $\R^{3}$, $\Sphere{3}$ can be regarded as attaching
two three-dimensional ball on two sides of two-sphere $\Sphere{2}$.
In this way, each circle in $\Sphere{3}$ corresponding to a single
point on the sphere in (a) can be represented as a circle in the three-dimensional
ball. Moreover, points in (a) are color coded corresponding to circles
in (b), e.g., one pole contains the red elliptical circle that would
become an infinite-radius circle by a slightly different way to represent
$\Sphere{3}$ in $\R^{3}$, and the opposite pole corresponds to the
large perfectly round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

@@

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Vec{a}$ and $\Vec{b}$ on $\Sphere{2}$,
the shortest rotation carrying the unit normal $\Hat{a}$ to the unit
normal $\Hat{b}$ is the SLERP (spherical linear interpolation) 
\begin{equation}
S(\Hat{a},\Hat{b},t)=\Hat{a}\,\frac{\sin((1-t)\omega)}{\sin\omega}\,+\,\Hat{b}\,\frac{\sin(t\omega)}{\sin\omega}\ ,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\omega$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Fig.~\ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.
